### 2016-07-25 finalText/proposedSol.tex
Paper ready, only missing introduction to architecture and results
Final Version of the Text and Short Paper

\subsection{Finite-precision representation system}
Before discussing any other details concerned with the actual hardware implementation, I will layout some of the design choices that were made regarding how the numbers involved in the calculations are represented. For that purpose, Section~\ref{sec:precbit} discusses the fundamentals of fixed-point representation systems, as well as the bitwidth and precision chosen for then number representation system of the network, and Section~\ref{sec:convrulesfp} states how the conversion between real and fixed-point numbers is performed. Finally, Section~\ref{sec:arithrulesfp} explains the special rules that fixed-point arithmetic imposes. The considerations made in this chapter can be found on~\cite{Yates13}, which is a good reference for fixed-point arithmetic theory.

\subsubsection{Precision bitwidth}\label{sec:precbit} 
Since we are dealing with real numbers, and I plan to make use of the DSP48 slices within the FPGA, I chose to use an 18-bit signed fixed-point system, with the sign information coded as 2's complement. Fixed-point systems are usually addressed in the $Qn.m$ form, where $n$ is the bitwidth of the of the integer part (excluding the sign bit) and $m$ is the bitwidth of the fractional part, and so the total bitwidth is $N=m+n+1$ to account for the sign bit. In this way, the value of the $i$-th position bit is $2^{i-m}$, and since this is a 2's complement system, the last bit is $-2^{n-1}$. In terms of range of representation, the maximum positive number that can be represented corresponds to all bits set to one, except for the last ($2^{n+m+1}-1$), but shifted by $m$ bits to the right to yield the correct real number (the decimal point is at the $m$-th bit),

@@

\begin{equation}\label{eq:minneg}
    \text{Small. Negative Number} = -\frac{2^{N-1}}{2^m} = -2^{N-1-m}
\end{equation}

Assuming that the smallest perturbation used in the training system will be $2^{-9}$, according to the previous Python experiments, our fractional part precision should be, at least, greater than this, so a sensible choice would be either $m=10$ or $m=11$. On the other hand, given that in the Python experiments I have attested that the weight values generally do not (and should not) exceed values around the first power of ten, there is no need for large values of $n$ (although it is advisable to be large enough to accommodate the intermediate calculations, avoiding overflow), and so we should choose to have as much precision as possible, so $m=11$. Since $18=n+m+1$, then $n = 17 - m = 6$ and the representation system to be used is $Q6.11$. According to Equations~\ref{eq:maxpos} and~\ref{eq:minneg}, the real values $x$ than can be represented with this choice of $n$ and $m$ are in the range

@@

\subsubsection{Fixed-point arithmetic rules}\label{sec:arithrulesfp}
The three main operations needed in my network design are \textbf{signed sums}, \textbf{signed multiplications} and \textbf{arithmetic shifts} (i.e. the ones that preserve the sign of the MSB) to implement divisions/multiplications by powers of two. In terms of signed sums, the rule is simple: both numbers need to be scaled to the same base, with their $m$'s being same, so that the decimal point is in the same place in both numbers ($n$, however, can be different, since that only means that one number is longer than the other, and the missing bits in the smaller one can be interpreted as zeros).

For signed multiplication, since both operands are in fixed-point $Qn.m$, and are thus scaled by $2^m$, we need to scale the result by multiplying it by $2^{-m}$ (or perform an arithmetic right shift of $m$ bits). This is because, if $a$ and $b$ the real numbers to be multiplied, we have that in fixed-point arithmetic, the result $c$ is

@@

\subsection{Non-linearity Calculator}\label{sec:nonlincalc}
In order to evaluate the non-linear activation functions $\sigma(x)$ and $\tanh(x)$, since there is no algorithm that can directly compute them, I had to find a suitable way to compute them accurately using a finite number of elementary operations, multiplications and additions, that can be performed efficiently by specially tailored blocks within the FPGA (DSP48 slices for multiplication, for instance). For that purpose, after using~\cite{Muller05} as a reference on elementary function approximation algorithms, I decided to use \textbf{Polynomial Approximations}, since evaluating a polynomial does not have high memory usage needs (as opposed to Table Methods, for instance) and, if the polynomial degree is sufficiently low, the number of multiplications needed is low enough to not pose a restriction both on resources (now DSP slices, and not memory) and in speed (number of number of clock cycles needed to output a result).

\subsubsection{Theoretical considerations on the approximation method}
The polynomial approximation methods aim to approximate some function $f(x)$ in an interval $\left[a, b\right]$ using a polynomial $p^*_n \in \mathcal{P}$ of degree $n$, in order to meet the optimization criteria chosen \textit{a priori}. This optimization criteria can be either the well-known \textbf{Least Squares Approximation} procedure, where we minimize the \emph{average quadratic error} $\left[f(x)-p^*(x)\right]^2$, or the \textbf{Least Maximum Approximation}, where we \emph{minimize} the \emph{maximum possible error}, also commonly called a \textit{minimax} approximation. Since we are operating in $Q6.11$ fixed-point arithmetic, we need to guarantee that the maximum approximation error does not exceed the precision limit of this representation, $2^{-11}$, and so a \textit{minimax} approach is desirable, since it guarantees that a given maximum error is not exceeded. Weierstrass's Theorem~\cite{Weierstrass1885}, from 1885, guarantees that there is always a polynomial that can approximate any continuous function f with error $\epsilon > 0$. Chebyshev also proved~\cite{Muller05} that, in a \textit{minimax} polynomial approximation of degree $n$, the minimum approximation error $\epsilon$ is achieved in at least $n+2$ points, and the sign of approximation error alternates from one interval to the other, and thus the error is not \emph{biased}. This leads to a linear system of $n+2$ equations whose $i$-th line is given by

@@

On one hand, the system of~\ref{eq:remezline} of Step 2 can be written in matricial notation as

\begin{equation}
\begin{bmatrix} 1 & x_0 & x_0^2 & \cdots & x_0^n & - 1 \\ 1 & x_1 & x_1^2 & \cdots & x_1^n & + 1 \\  &  &  & \vdots & & \\ 1 & x_{n+1} & x_{n+1}^2 & \cdots & x_{n+1}^n & (-1)^{n+1} \end{bmatrix}
\begin{bmatrix} p_0 \\ p_1 \\ p_2 \\ \vdots \\ p_n \\ \epsilon \end{bmatrix} = \begin{bmatrix} f(x_0) \\ f(x_1) \\ \vdots \\ f(x_{n+1}) \end{bmatrix}
\end{equation}
and therefore the solution vector $p = \left[p_0 \, p_1 \, p_2 \, \cdots \, p_n \, \epsilon \right]$ for Step 2 of Remez's Algorithm is simply given by $p = A^{-1}b$, where $A$ is the matrix on the left-hand side of the equation and $b$ is the vector on the right-hand side. The Python implementation of this algorithm is presented in Listing~\ref{lst:remezpy} of Appendix~\ref{ap1:remez-pyth}.

Instead of using a \emph{single} polynomial of higher degree ($n \geq 3$) for the whole domain, I chose to partition the domain of the activation functions in \textbf{6 intervals}, and approximate each of those intervals using polynomials of degree $n=2$. This proved to yield a lower overall approximation error, as expected (the interval on which to perform the approximation is smaller). Also, since both the $\sigma(x)$ and $\tanh(x)$ have horizontal asymptotes in $\{0,1\}$ and $\{-1,1\}$ respectively, the far-left and far-right intervals do not need a polynomial approximation, and can be assigned a constant value equal to the corresponding asymptote. The resulting \textit{minimax} approximation polynomials yielded by running the code in Listing~\ref{lst:remezpy} are

\begin{equation}\label{eq:coefs_sigm}
\hat{\sigma(x)} = \left\{ 
\begin{array}{lc} 
0 & x  \leq -6 \\
0.20323428 + 0.0717631x + 0.00642858x^2 & -6 \leq x \leq -3 \\
0.50195831 + 0.27269294x + 0.04059181x^2 & -3 \leq  x \leq 0 \\
0.49805785 + 0.27266221x - 0.04058115x^2 &  0 \leq  x \leq 3 \\
0.7967568 + 0.07175359x - 0.00642671x^2 & 3 \leq  x \leq 6 \\
1 & x > 6
\end{array}
\right.
\end{equation}
for the sigmoid function and

\begin{equation}\label{eq:coefs_tanh}
\hat{\tanh(x)} = \left\{ 
\begin{array}{lc} 
-1 & x  \leq -3 \\
-0.39814608 + 0.46527859x + 0.09007576x^2 & -3 \leq x \leq -1 \\
0.0031444 + 1.08381219x + 0.31592922x^2 & -1 \leq  x \leq 0 \\
-0.00349517 + 1.08538355x -0.31676793x^2 &  0 \leq  x \leq 1 \\
0.39878032 + 0.46509003x - 0.09013554x^2 & 1 \leq  x \leq 3 \\
1 & x > 3
\end{array}
\right.
\end{equation}
for the hyperbolic tangent function. These constants were converted to $Qn.m$ using the rules in~\ref{sec:convrulesfp}, and embedded in the HDL model.

\subsubsection{Hardware Implementation}
The hardware module that implements these non-linearities is, essentially, a 2nd degree polynomial calculator, where the coefficients of the polynomial are chosen accordingly with the value of the input $x$. This last functionality
is implemented using a simple multiplexer that loads the signals $p_0$, $p_1$ and $p_2$ with the coefficients of Equations~\ref{eq:coefs_sigm} and~\ref{eq:coefs_tanh} based on the value of the input operand $x$. As for the polynomial calculator is concerned, although we could use a full-pipelined evaluator, that would require two DSP slices -- to perform the two simultaneous multiplications -- but the DSP slices will be heavily used in the matrix-vector calculators, so it is advisable to save them for that purpose. A simpler approach is to note that, according to Horner's Rule, we get

\begin{equation}\label{eq:factorPol}
p(x) = p_0 + p_1x + p_2x^2 = p_0 + x(p_1 + xp_2)
\end{equation}
where we can note that this operation can be divided in a two-step procedure of a simultaneous multiplication of the operand by a constant, and a subsequent addition of another constant: first, by multiplying the operand by $p_2$ and summing $p_1$, and then by multiplying the operand by this last result and summing $p_0$. The block diagram of the hardware implementation of the non-linearity calculator is presented in Figure~\ref{fig:nonlin}. Also, in Figure~\ref{fig:nonlin-out}, the output of the Verilog module that implements this design is compared with the actual output (I used Python3's Numpy as reference) for both activation functions.

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{figures/non-lin.eps}
    \caption[Block Diagram of the Non-linearity Calculator]{Block Diagram of the Non-linearity Calculator using a single multiplication. The multiplexer state is controlled by the flip-flip and sum block, that change state every clock cycle. When reset is applied, the selector is set to zero.}
    \label{fig:nonlin}
\end{figure}

@@

\subsection{Matrix-vector Dot Product Unit}\label{sec:dotprod_sec}
From the set of Equations~\ref{eq:equationsLSTM}, we see that the weight matrices $\mb{W}_*$ and $\mb{R}_*$ are multiplied by the input vector $\mb{x}$ and the layer output vector $\mb{y}$, respectively. This way, we need a HDL block that implements matrix-vector multiplication, and that block must be parameterized in order to accommodate different matrices/vectors of various sizes: note that $x$ has length $M$ (the number of inputs to the layer), and so $\mb{W}_*$ has size $N\times M$, while $y$ has length $N$ (the number of neurons in the layer), and so $\mb{R}_*$ now has size $N\times N$. Plus, if we need a layer with different parameters, either in terms of number of inputs or number of neurons, it advisable to only change the respective parameter at the synthesis stage instead of having to redesign the whole block for that particular size.

The matrix-vector dot product of a matrix $A$ of size $N \times M$ by a vector $x$ of size $M$, if performed in a linear non-parallel way, can be described in terms of Algorithm~\ref{matvec-alg}

\begin{algorithm}
\begin{algorithmic}
\For {$i$ = $1:N$}
    \For {$j$ = $1:M$}
    \State $y_i := y_i + A_{ij} \cdot x_j$
    \EndFor
\EndFor
\end{algorithmic}
\caption{Matrix-vector multiplication of a matrix}
\label{matvec-alg}
\end{algorithm}
This operation has a computational complexity of $O(n^2)$. It can be seen that each of the $i$-th component of the output vector $y$ can be calculated \textbf{in parallel}, each only requiring the corresponding $i$-th line from the matrix. If we follow this approach, matrix-vector multiplication can now be performed in \textbf{linear time}, which is one of the great advantages of custom-tailored hardware solutions.

Although this solution only requires one multiplication per row of the input matrix (i.e. $N$ multiplications), if the row size is large, we may run out of resources in the FPGA; therefore, some sort of \textit{resource multiplexing} strategy must be used to ensure the flexibility of the solution to accomodate networks of larger dimensions. The solution I have found for this issue was to \emph{share} the multiplication slice between rows of the matrix: in a direct implementation of Algorithm~\ref{matvec-alg}, each multiplication slice was responsible for producing the $i$-th element of the output vector $y$ (of size $N$), therefore the final result for the vector would be ready in $M$ clock cycles (i.e. the number of columns); now, if we define a parameter $K_G = \frac{\text{Number of rows}}{\text{Number of multipliers}}$ -- the number of rows that share the same multiplier -- the same multiplier is responsible for producing several $i$-th elements of the output vector, in consecutive time slots of $M$ clock cycles. Suppose that we have a $8\times2$ matrix, and that $K_G = 2$; in this way, we would have 4 multipliers, and the output vector elements $y_0$, $y_2$, $y_4$ and $y_6$ would be ready after $M=2$ clock cycles, and the remaining -- $y_1$, $y_3$, $y_5$ and $y_7$ --  are ready after another two clock cycles, that is $2M = 4$ clock cycles after the calculations began.

Figures-\ref{fig:mem-arrayprod} and~\ref{fig:array-prod} depict a diagram of the memory access for the Matrix, and the row multiplication and units within the module, respectively, where I have set $K_G = 4$, for the same matrix and vector sizes as before. Note that in this situation, we would only have 2 multipliers, and the module would be composed of two multiplication units, such as those in Figure~\ref{fig:array-prod}, that work in parallel: they address a particular column using the signal \verb+colSel+, which is used by the RAM module to output the corresponding column of the matrix (in regard to the input vector, obviously this signal selects only a single position), depicted in Figure~\ref{fig:mem-arrayprod}. The dark shaded part of the memory is used by the first multiplier, and the light shaded is used by the other, in parallel for a fixed \verb+rowMux+ -- this signal is produced by the control unit of the module, and essentially operates the left Mux and right Demux of Figure~\ref{fig:array-prod} that allows to choose the proper position of the weight column and to write to the correct output register, respectively. In this example, for \verb+rowMux+=0, the control unit increments \verb+colSel+ from 0 to $M$, and thus evaluating $y_0$ and $y_4$. After this, \verb+rowMux+ is incremented to 1, and the process repeats until \verb+rowMux+ reaches $K_G-1$. Therefore, we have the correct result vector in a time proportional to $K_G \cdot M$: since the memory only outputs the appropriate column in the \emph{next clock cycle}, and that this module is pipelined both at the input and at the output in order to increase the maximum clock frequency, we should add a 3 clock cycle overhead to the previous estimate.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/array-prod.eps}
    \caption[The $i$-th row multiplication unit of the Module]{The $i$-th row multiplication unit of the Module, where rowMux and colSel are internal signals produced by the control unit of the module. The flip-flop accumulates the sum, and the output demux selects the appropriate memory position on where to store this value, within the slot attributed to this multiplication, from $i\cdot K_G$ to $i\cdot K_G + \left[K_G-1\right]$}
    \label{fig:array-prod}
\end{figure}

\subsection{Dual-port weight RAM}
Even though I have chosen to keep the bias weight vectors in normal registers and let the synthesis tool decide how to place them in the FPGA, for the weight matrices, the case is different. Since the network size can be large, it is good to make sure that they are placed in the RAMs available in the FPGA. In Section~\ref{sec:dotprod_sec}, we see that, because of the architecture of the matrix-vector multiplier, it is convenient to have a column by column access to the weight matrix, and therefore this block outputs a given column in the \verb+rowOut+ output port terminal, at the \emph{negative edge} of the clock, selected by the value present in the input \verb+addressOut+ at the immediately previous \emph{positive edge} of the clock. As far as \textbf{writing} to the memory is concerned, the process is identical but now the column of weights at the input \verb+rowIn+ is sampled at \emph{negative edge} of the clock, and is placed at the column specified by the input \verb+addressIn+ at the previous \emph{positive edge} of the clock.

@@

The Verilog coding followed the Verilog Coding Guidelines in Xilinx's \href{http://www.xilinx.com/support/documentation/sw_manuals/xilinx2015_2/ug901-vivado-synthesis.pdf}{UG901} that, in page 50, recommends that the \verb+RAM_STYLE+ parameter be set to \verb+block+. This can be done by adding the follwing compiler directive before the register definition, as follows

@@

After performing synthesis, I have noticed that using block RAM led to a slightly poorer performance, so I have settled with the use of \emph{Distributed RAM}, storing the weight matrices in LUTRAM.

\subsection{Gate Module}\label{sec:gatemod} 
The Gate modules are responsible for producing the internal signal vectors for $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$. If we note that, according to~\cite{Greff15}, the removal of peepholes (the signals $\mb{p}_*$) does not compromise significantly the performance of the network~\ref{sec:struct_lstm}, we can ommit them in order to simplify the Gate Module and reduce the usage of DSP slices. This way, a Gate module needs to perform three tasks

>>>>

\subsection{Finite-precision representation system}
Before discussing any other details concerned with the actual hardware implementation, some of the design choices that were
made regarding how the numbers involved in the calculations are represented are laid out. For that purpose, Section~\ref{sec:precbit} discusses the fundamentals of fixed-point
representation systems, as well as the bitwidth and precision chosen for the number representation system of the network, and Section~\ref{sec:convrulesfp} states
how the conversion between real and fixed-point numbers is performed. Finally, Section~\ref{sec:arithrulesfp} explains the special rules that fixed-point arithmetic
imposes. The considerations made in this chapter can be found on~\cite{Yates13}, which is a good reference for fixed-point arithmetic theory.

\subsubsection{Precision bitwidth}\label{sec:precbit}
Since we are dealing with real numbers, and the plan is to make use of the DSP48 slices within the FPGA, an 18-bit signed fixed-point system was chosen, with the sign
information coded as 2's complement. Fixed-point systems are usually addressed in the $Qn.m$ form, where $n$ is the bitwidth of the of the integer part (excluding
the sign bit) and $m$ is the bitwidth of the fractional part, and so the total bitwidth is $N=m+n+1$ to account for the sign bit. In this way, the value of
the $i$-th position bit is $2^{i-m}$, and since this is a 2's complement system, the last bit is worth $-2^{n-1}$. In terms of range of representation, the maximum positive
number that can be represented corresponds to all bits set to one, except for the last ($2^{n+m+1}-1$), but shifted by $m$ bits to the right to yield the correct real
number (the decimal point is at the $m$-th bit), i.e.

@@

\begin{equation}\label{eq:minneg}
    \text{Small. Negative Number} = -\frac{2^{N-1}}{2^m} = -2^{N-1-m}.
\end{equation}

Assuming that the smallest perturbation used in the training system will be $2^{-9}$, according to the previous Python experiments, the fractional part precision should be,
at least, greater than this, so a sensible choice would be either $m=10$ or $m=11$. On the other hand, given that, in the Python experiments, it was attested that the
weight values generally do not (and should not) exceed values around the first power of ten, there is no need for large values of $n$ (although it is advisable to be large
enough to accommodate the intermediate calculations, avoiding overflow), and so we should choose to have as much precision as possible, so $m$ was set to 11. Since $18=n+m+1$, then
$n = 17 - m = 6$ and the representation system to be used is $Q6.11$. According to Equations~\ref{eq:maxpos} and~\ref{eq:minneg}, the real values $x$ than can be represented
with this choice of $n$ and $m$ are in the range

@@

\subsubsection{Fixed-point arithmetic rules}\label{sec:arithrulesfp}
The three main operations needed in my network design are \textbf{signed sums}, \textbf{signed multiplications} and \textbf{arithmetic shifts} (i.e. the ones that preserve the sign of the MSB)
to implement divisions/multiplications by powers of two. In terms of signed sums, the rule is simple: both numbers need to be scaled to the same base, with their $m$'s being same,
so that the decimal point is in the same place in both numbers ($n$, however, can be different, since that only means that one number is longer than the other, and the missing bits
in the smaller one are interpreted as zeros).

For signed multiplication, since both operands are in fixed-point $Qn.m$, and are thus scaled by $2^m$, we need to scale the result by multiplying it
by $2^{-m}$ (or perform an arithmetic right shift of $m$ bits). This is because, if $a$ and $b$ are the real numbers to be multiplied, and $c$ is the
result, we have that, in fixed-point arithmetic, the result is

@@

\subsection{Activation Function Calculator}\label{sec:nonlincalc}
In order to evaluate the non-linear activation functions $\sigma(x)$ and $\tanh(x)$, since there is no algorithm that can directly compute them,
a suitable way to compute them accurately had to be found, using a finite number of elementary operations, multiplications and additions, that can be performed
efficiently by specially tailored blocks within the FPGA (DSP48 slices for multiplication, for instance). For that purpose, after using~\cite{Muller05} as a
reference on elementary function approximation algorithms, \textbf{Polynomial Approximations} were used, since evaluating a polynomial does not have high memory usage
needs (as opposed to Table Methods, for instance) and, if the polynomial degree is sufficiently low, the number of multiplications needed is low enough to not pose a restriction
both on resources (now DSP slices, and not memory) and in speed (number of number of clock cycles needed to output a result).

\subsubsection{Theoretical considerations on the approximation method}
The polynomial approximation methods aim to approximate some function $f(x)$ in an interval $\left[a, b\right]$ using a polynomial $p^*_n \in \mathcal{P}$ of
degree $n$, in order to meet the optimization criteria chosen \textit{a priori}. This optimization criteria can be either the
well-known \textbf{Least Squares Approximation} procedure, where we minimize the \emph{average quadratic error} $\left[f(x)-p^*(x)\right]^2$, or
the \textbf{Least Maximum Approximation}, where we \emph{minimize} the \emph{maximum possible error}, also commonly called a \textit{minimax} approximation.
Since we are operating in $Q6.11$ fixed-point arithmetic, we need to guarantee that the maximum approximation error is close to the precision limit of this
representation, $2^{-11}$, and so a \textit{minimax} approach is desirable, since it guarantees that a given maximum error is not exceeded. Weierstrass's
Theorem guarantees that there is always a polynomial that can approximate any continuous function f with error $\epsilon > 0$.
Chebyshev also proved~\cite{Muller05} that, in a \textit{minimax} polynomial approximation of degree $n$, the minimum approximation error $\epsilon$ is achieved in
at least $n+2$ points, and the sign of the approximation error alternates from one interval to the other, meaning that the error is not \emph{biased} and has zero average.
This leads to a linear system of $n+2$ equations, whose $i$-th line is given by

@@

On one hand, the system of~\ref{eq:remezline} of Step 2 can be written in matrix notation as

\begin{equation}
\begin{bmatrix} 1 & x_0 & x_0^2 & \cdots & x_0^n & - 1 \\ 1 & x_1 & x_1^2 & \cdots & x_1^n & + 1 \\  &  &  & \vdots & & \\ 1 & x_{n+1} & x_{n+1}^2 & \cdots & x_{n+1}^n & (-1)^{n+1} \end{bmatrix}
\begin{bmatrix} p_0 \\ p_1 \\ p_2 \\ \vdots \\ p_n \\ \epsilon \end{bmatrix} = \begin{bmatrix} f(x_0) \\ f(x_1) \\ \vdots \\ f(x_{n+1}) \end{bmatrix}
\end{equation}
and therefore the solution vector $p = \left[p_0 \; p_1 \; p_2 \; \cdots \; p_n \; \epsilon \right]$ for Step 2 of Remez's Algorithm is simply given by $p = A^{-1}b$, where $A$ is the matrix on the left-hand side of the equation and $b$ is the vector on the right-hand side. The Python implementation of this algorithm is presented in Listing~\ref{lst:remezpy} of Appendix~\ref{ap1:remez-pyth}.

Instead of using a \emph{single} polynomial of higher degree ($n \geq 3$) for the whole domain, the domain of the activation functions was partitioned
in \textbf{6 intervals}, and polynomials of degree $n=2$ were used to approximate the function in each of those intervals. This proved to yield a lower overall approximation error,
as expected (the interval on which to perform the approximation is smaller). Also, since both $\sigma(x)$ and $\tanh(x)$ have horizontal asymptotes
in $\{0,1\}$ and $\{-1,1\}$ respectively, the far-left and far-right intervals do not need a polynomial approximation, and can be assigned a constant value
equal to the corresponding asymptote. The resulting \textit{minimax} approximation polynomials yielded by running the code in Listing~\ref{lst:remezpy} are

\begin{equation}\label{eq:coefs_sigm}
\sigma(x) \approx \left\{
\begin{array}{lc}
0 & x  \leq -6 \\
0.20323428 + 0.0717631x + 0.00642858x^2 & -6 \leq x \leq -3 \\
0.50195831 + 0.27269294x + 0.04059181x^2 & -3 \leq  x \leq 0 \\
0.49805785 + 0.27266221x - 0.04058115x^2 &  0 \leq  x \leq 3 \\
0.7967568 + 0.07175359x - 0.00642671x^2 & 3 \leq  x \leq 6 \\
1 & x > 6
\end{array}
\right.
\end{equation}
for the sigmoid function and

\begin{equation}\label{eq:coefs_tanh}
\tanh(x) \approx \left\{
\begin{array}{lc}
-1 & x  \leq -3 \\
-0.39814608 + 0.46527859x + 0.09007576x^2 & -3 \leq x \leq -1 \\
0.0031444 + 1.08381219x + 0.31592922x^2 & -1 \leq  x \leq 0 \\
-0.00349517 + 1.08538355x -0.31676793x^2 &  0 \leq  x \leq 1 \\
0.39878032 + 0.46509003x - 0.09013554x^2 & 1 \leq  x \leq 3 \\
1 & x > 3
\end{array}
\right.
\end{equation}
for the hyperbolic tangent function. These constants were converted to $Qn.m$ using the rules in~\ref{sec:convrulesfp}, and embedded in the HDL model.

\subsubsection{Hardware Implementation}
The hardware module that implements these activation functions is, essentially, a 2nd degree polynomial calculator, where the coefficients of the polynomial are chosen accordingly with the value of the input $x$. This last functionality
is implemented using a simple multiplexer that loads the signals $p_0$, $p_1$ and $p_2$ with the coefficients of Equations~\ref{eq:coefs_sigm} and~\ref{eq:coefs_tanh} based on the value of the input operand $x$. As for the polynomial calculator is concerned, although we could use a full-pipelined evaluator, that would require two DSP slices -- to perform the two simultaneous multiplications -- but the DSP slices will be heavily used in the matrix-vector calculators, so it is advisable to save them for that purpose. A simpler approach is to note that, according to Horner's Rule, we get

\begin{equation}\label{eq:factorPol}
p(x) = p_0 + p_1x + p_2x^2 = p_0 + x(p_1 + xp_2)
\end{equation}
where we can note that this operation can be divided in a two-step procedure of a simultaneous multiplication of the operand by a constant,
and a subsequent addition of another constant: first, by multiplying the operand by $p_2$ and summing $p_1$, and then by multiplying the operand
by this last result and summing $p_0$. The block diagram of the hardware implementation of the non-linearity calculator is presented in Figure~\ref{fig:nonlin}.
Also, in Figure~\ref{fig:nonlin-out}, the output of the Verilog module that implements this design is compared with the actual output (Python3's Numpy was used as reference)
for both activation functions. The maximum error for the approximation of the $\sigma(x)$ was of 0.001408 and for the $\tanh(x)$ was 0.0121.

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{figures/non-lin.eps}
    \caption[Block Diagram of the Activation Function Calculator]{Block Diagram of the Activation Function Calculator using a single multiplication. The multiplexer state is controlled by the flip-flip and sum block, that change state every clock cycle. When reset is applied, the selector is set to zero.}
    \label{fig:nonlin}
\end{figure}

@@

\subsection{Matrix-vector Dot Product Unit}\label{sec:dotprod_sec}
From the set of Equations~\ref{eq:equationsLSTM}, we see that the weight matrices $\mb{W}_*$ and $\mb{R}_*$ are multiplied by the input
vector $\mb{x}$ and the layer output vector $\mb{y}$, respectively. This way, we need an HDL block that implements matrix-vector multiplication,
and the description of that block must be parameterized in order to accommodate different matrices/vectors of various sizes: note that $x$ has length $M$ (the number
of inputs to the layer), and so $\mb{W}_*$ has size $N\times M$, while $y$ has length $N$ (the number of neurons in the layer), and so $\mb{R}_*$
now has size $N\times N$. Plus, a layer with different parameters is needed, either in terms of number of inputs or number of neurons, it is advisable
to only change the respective parameter at the synthesis stage, instead of having to redesign the whole block for that particular size.

The matrix-vector dot product of a matrix $A$ of size $N \times M$ with a vector $\mb{x}$ of size $M$, if performed in a linear non-parallel way, can be described in terms of Algorithm~\ref{matvec-alg}

\begin{algorithm}
\begin{algorithmic}
\For {$i$ = $1:N$}
    \For {$j$ = $1:M$}
    \State $y_i := y_i + A_{ij} \cdot x_j$
    \EndFor
\EndFor
\end{algorithmic}
\caption{Matrix-vector multiplication of a matrix}
\label{matvec-alg}
\end{algorithm}
This operation has a computational complexity of $O(n^2)$. It can be seen that each of the $i$-th components of the output vector $\mb{y}$ can be calculated \textbf{in parallel},
each only requiring the corresponding $i$-th line from the matrix. Following this approach, matrix-vector multiplication can now be performed in \textbf{linear time}, which
is one of the great advantages of custom-tailored hardware solutions.

Although this solution only requires one multiplication per row of the input matrix (i.e. $N$ multiplications), if the row size is large, we may quickly run out of resources in the FPGA
as $N$ increases; therefore, some sort of \textit{resource multiplexing} strategy must be used to ensure the flexibility of the solution to accommodate networks of larger dimensions.
The solution found for this issue was to \emph{share} the multiplication slice between rows of the matrix: in a direct implementation of Algorithm~\ref{matvec-alg}, each
multiplication slice was responsible for producing the $i$-th element of the output vector $\mb{y}$ (of size $N$), therefore the final result for the vector would be ready in $M$
clock cycles (i.e. the number of columns); now, if defining a parameter $K_G = \frac{\text{Number of rows}}{\text{Number of multipliers}}$, the number of rows that share the
same multiplier, each multiplier is responsible for producing several $i$-th elements of the output vector, in consecutive time slots of $M$ clock cycles. Suppose that we
have a $8\times2$ matrix, and that $K_G = 2$; in this way, we would have 4 multipliers, and the output vector elements $y_0$, $y_2$, $y_4$ and $y_6$ would be ready after $M=2$
clock cycles, and the remaining -- $y_1$, $y_3$, $y_5$ and $y_7$ --  are ready after another two clock cycles, that is $2M = 4$ clock cycles after the calculations began.

Figures-\ref{fig:mem-arrayprod} and~\ref{fig:array-prod} depict a diagram of the memory access for the Matrix, and the row multiplication units within the module, respectively,
where $K_G = 4$ was set, for the same matrix and vector sizes as before. Note that in this situation there would be only 2 multipliers, and therefore the module would be composed of two
multiplication units, such as those in Figure~\ref{fig:array-prod}, that work in parallel: they address a particular column using the signal \verb+colSel+, which is used by the RAM
module to output the corresponding column of the matrix (in regard to the input vector, obviously this signal selects only a single position), depicted in Figure~\ref{fig:mem-arrayprod}.
The dark shaded part of the memory is used by the first multiplier, and the light shaded is used by the other, in parallel for a fixed \verb+rowMux+ -- this signal is produced by the
control unit of the module, and essentially operates the left Mux and right Demux of Figure~\ref{fig:array-prod} that allows to choose the proper position of the weight column and to
write to the correct output register, respectively. In this example, for \verb+rowMux+=0, the control unit increments \verb+colSel+ from 0 to $M$, and thus evaluating $y_0$ and $y_4$.
After this, \verb+rowMux+ is incremented to 1, and the process repeats until \verb+rowMux+ reaches $K_G-1$. Therefore, we have the correct result vector in a time proportional to $K_G \cdot M$:
since the memory only outputs the appropriate column in the \emph{next clock cycle}, and that this module is pipelined both at the input and at the output in order to increase the maximum clock
frequency, we should add a 3 clock cycle overhead to the previous estimate.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/array-prod.eps}
    \caption[The $i$-th row multiplication unit of the Module]{The $i$-th row multiplication unit of the Module, where rowMux and colSel are internal signals produced by the control unit of the module. The flip-flop accumulates the sum, and the output Demux selects the appropriate memory position on where to store this value, within the slot attributed to this multiplication, from $i\cdot K_G$ to $i\cdot K_G + \left[K_G-1\right]$}
    \label{fig:array-prod}
\end{figure}

\subsection{Dual-port weight RAM}
Even though bias weight vectors, with size $N\times 1$, were kept as normal registers and the synthesis tool was left to decide how to place them in the FPGA,
for the weight matrices the case is different. Since the network size can be large, it is good to make sure that they are placed in the RAMs available
in the FPGA. In Section~\ref{sec:dotprod_sec} we see that, because of the architecture of the matrix-vector multiplier, it is convenient to have a column
by column access to the weight matrix, and therefore this block was designed to output a given column in the \verb+rowOut+ output port terminal, at the \emph{positive edge} of
the clock, selected by the value present in the input \verb+addressOut+ at the immediately previous \emph{positive edge} of the clock. As far as \textbf{writing} to
the memory is concerned, the process is identical but now the column of weights at the input \verb+rowIn+ is sampled at \emph{positive edge} of the clock, and is placed
at the column specified by the input \verb+addressIn+ at the previous \emph{positive edge} of the clock.

@@

The Verilog coding followed the Verilog Coding Guidelines in Xilinx's \href{http://www.xilinx.com/support/documentation/sw_manuals/xilinx2015_2/ug901-vivado-synthesis.pdf}{UG901 Document} that,
in page 50, recommends that the \verb+RAM_STYLE+ parameter be set to \verb+block+. This can be done by adding the following compiler directive before the register definition, as follows

@@

After performing synthesis, it was noticed that using block RAM led to a slightly poorer speed performance, so I have settled with the use of \emph{Distributed RAM}, storing the weight matrices in LUTRAM.

\subsection{Gate Module}\label{sec:gatemod}
The Gate modules are responsible for producing the internal signal vectors for $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$. If we note that, according
to~\cite{Greff15}, the removal of peepholes (the signals $\mb{p}_*$) does not compromise significantly the performance of the network~\ref{sec:struct_lstm}, they were omitted
in order to simplify the Gate Module and reduce the usage of DSP slices. This way, a Gate module needs to perform three tasks
